---
title: "Simulation of tree-indexed random walks"
output: html_document
date: "2024-07-01"
---

```{r echo=FALSE, eval = FALSE}
library(evd)
```

#### define a, L, psi
```{r echo=FALSE, eval = FALSE}

# Example 2 of Gantert
psi <- function(n, r = 0.5) {
  
  psi <- ifelse(n > 1, ( r*n / log(n) )^(1 / r), 1)
  
  return(psi)
}

L <- function(n, r) {
  L <- log(n)
  
  return(L)
}

# assume a = 1 (for simplification)
a <- function(n) {
  return(1)
}

r = 0.5


```

```{r echo=FALSE, eval = FALSE}
mu = 0
sd = 1

loc = 0
scale = 1
shape = 1

p = 0.75
```


#### define the support
```{r echo=FALSE, eval = FALSE}
n_min = 10^1
n_max = 10^3

support <- seq(from = n_min, to = n_max, by = 1)
```


#### plot psi vs. n
```{r echo=FALSE, eval = TRUE}
plot(support, psi(support))
lines(support, support)
```


#### semiexponential tails
```{r echo=FALSE, eval = TRUE}
n_min_tail = 10^4
n_max_tail = 10^6
N = 10^4

tail_support <- seq(from = n_min_tail, to = n_max_tail, by = (n_max_tail - n_min_tail) / N)

upper_tail_dist <- function(t, t_min, r, a, L) {
  
  prob = a(t) * exp(-L(t, r = r) * t^r) * (t >= t_min)
    
  return(prob)
}

distribution_stretched_exp_tail <- function(t, t_min, p = 0.75, r = 0.5, a = 1, L, lower_tail_dist, mu = 0, sd = 1){
  
  prob_upper_tail <- (1 - p) + p * (1 - upper_tail_dist(t = t, t_min = t_min, r = r, a = a, L = L))
  
  if (lower_tail_dist == "normal") {
      prob_lower_tail <- (1 - p) * pnorm(t, mean = mu, sd = sd)
  }
  
  prob <- ifelse(t > t_min, prob_upper_tail, ifelse(t == t_min, 1 - p, prob_lower_tail))
  
  return(prob)
}

t_min = 2
distribution_stretched_exp_tail(t = t_min, t_min, r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd)
```

#### fréchet upper tail
```{r echo=FALSE, eval = TRUE}
distribution_frechet_tail <- function(t, t_min, p = 0.75, loc = 0, scale = 1, shape = 1, lower_tail_dist = "normal", mu = 0, sd = 1) {
  
  prob_upper_tail <- (1 - p) + p * pfrechet(q = t, loc = loc, scale = scale, shape = shape, lower.tail = TRUE) 
  
  if (lower_tail_dist == "normal") {
      prob_lower_tail <- (1 - p) * pnorm(t, mean = mu, sd = sd)
  }
  
  prob <- ifelse(t > t_min, prob_upper_tail, ifelse(t == t_min, 1 - p, prob_lower_tail))
  
  return(prob)
}

```

#### plot cdf
```{r echo=FALSE, eval = TRUE, warning = FALSE}
n_min = -5
n_max = 5

support <- seq(from = n_min, to = n_max, by = 0.05)

dist_stretched_exp_tail <- distribution_stretched_exp_tail(
  t = support, t_min = 2, r = 0.5, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

dist_frechet <- distribution_frechet_tail(t = support, t_min, p = p, loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd)

plot(support, dist_stretched_exp_tail, type = "s", ylab = "distribution", xlab = "")
lines(support, pnorm(support, mean = 0, sd = 1), col = "blue")
lines(support, dist_frechet, type = "s", col = "orange")
lines(support, rep(x = 1 - p, times = length(support)), col = "lightgray", lty = 2)

```

#### approximate the density of the stretched exponential distribution
```{r echo=FALSE, eval = TRUE}
n_support <- 10^3
n_min <- -10 + 1 / n_support
n_max <- 10
support_density <- seq(from = n_min, to = n_max, by = 1 / n_support)

## use a centered difference approximation to estimate the density
density_approx <- function(
  t, h = 1e-5, t_min = 2, p = 0.75, upper_tail_dist = "stretched", r = 0.5, a = function() {1}, L = log, loc = 0, scale = 1, shape = 1, lower_tail_dist = "normal", mu = 0, sd = 1
) {
  
  if (h >= min(diff(support_density))) {
    warning("The numerical differentiation will not work at the discontinuity t_min.")
  }
  
  if (upper_tail_dist == "stretched") {
    density_approx <- ( 
      distribution_stretched_exp_tail(
        t = t+h, t_min = t_min, p = p, r = r, a = a, L = L, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      ) - distribution_stretched_exp_tail(
        t = t-h, t_min = t_min, p = p, r = r, a = a, L = L, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      ) 
    ) / (2 * h)
  } else if (upper_tail_dist == "frechet") {
    density_approx <- ( 
      distribution_frechet_tail(
        t = t+h, t_min, p = p, loc = loc, scale = scale, shape = shape, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      ) - distribution_frechet_tail(
          t = t-h, t_min, p = p, loc = loc, scale = scale, shape = shape, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      )
    ) / (2 * h)
  }
  
  ## density is not defined at the discontinuity t = t_min
  density_approx <- ifelse(t != t_min, density_approx, NA_real_)
  
  return(density_approx)
}

```

#### analytical representation of density
```{r echo=FALSE, eval = TRUE}
density_analytical <- function(
    t, t_min = 2, p = 0.75, upper_tail_dist = "stretched", r = 0.5, a = function() {1}, L = log, loc = 0, scale = 1, shape = 1, lower_tail_dist = "normal", mu = 0, sd = 1
) {
  
  if (lower_tail_dist == "normal") {
    density_lower_tail <- (1 - p) * dnorm(t, mean = mu, sd = sd)
  }
  
  if (upper_tail_dist == "stretched") {
    density_upper_tail <- ifelse(t > 0, p * ( a(t) * t^(r-1) * exp(-L(t, r = r)*t^r) * (r * log(t) + 1) ), NA_real_)
  } else if (upper_tail_dist == "frechet") {
      density_upper_tail <- ifelse(t > 0, p * dfrechet(t, loc = loc, scale = scale, shape = shape), NA_real_)
  }
  
  density <- ifelse(t < t_min, density_lower_tail, density_upper_tail)
  
  return(density)
}

```

#### plot pdf
```{r echo=FALSE, eval = TRUE, warning = FALSE}

h = min(1e-7, min(diff(support_density)) / 2)

dist_stretched_exp_tail_approx <- distribution_stretched_exp_tail(
  t = support_density, t_min = 2, r = 0.5, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_stretched_exp_tail_approx <- density_approx(
  t = support_density, h = h, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_stretched_exp_tail_analytical <- density_analytical(
  t = support_density, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_frechet_tail_approx <- density_approx(
  t = support_density, h = h, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_frechet_tail_analytical <- density_analytical(
  t = support_density, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_norm <- dnorm(support_density, mean = mu, sd = sd)

y_max = 0.075 + max(
  max(density_stretched_exp_tail, na.rm = TRUE), 
  max(density_frechet, na.rm = TRUE), 
  max(density_norm)
)

plot(support_density, density_stretched_exp_tail_approx, type = "s", ylim = c(0, y_max), ylab = "density", xlab = "")
lines(support_density, density_stretched_exp_tail_analytical, col = "blue", lty = 2)

plot(support_density, density_frechet_tail_approx, type = "s", ylim = c(0, y_max), ylab = "density", xlab = "")
lines(support_density, density_frechet_tail_analytical, col = "red", lty = 2)

```




#### Acceptance-Rejection Algorithm

#### using approximated densities
```{r echo=FALSE, eval = TRUE}

n_support_approx <- 10^3
n_min_approx <- t_min - 1 / n_support
n_max_approx <- 10^3
support_density_approx <- seq(from = n_min_approx, to = n_max_approx, by = 1 / n_support_approx)


## scaling factor (lower tail is the same for both distributions)
C_lower_tail <- 1

## the frechet distribution has a heavier tail than the semiexponential case
C_upper_tail <- max(
  density_approx(
  t = support_density_approx, h = min(1e-7, min(diff(support_density_approx)) / 2), t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
) / density_approx(
      t = support_density_approx, h = min(1e-7, min(diff(support_density_approx)) / 2), t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
  ), na.rm = TRUE
)

C <- max(C_lower_tail, C_upper_tail)

```

#### Accept-Reject using the numerical approximation of densities
```{r echo=FALSE, eval = TRUE, warning = FALSE}
X <- c()
k = 1
h = 1e-5

set.seed(42)

while (k < 5 * 10^3) {
  B <- rbinom(1, 1, prob = p)
  y_norm <- min(rnorm(n = 1, mean = mu, sd = 1), t_min - 2*h)
  
  ## sample X ~ Frechet, conditional on X > 2
  u_uniform_for_frechet <- runif(1, min = pfrechet(2, loc = loc, scale = scale, shape = shape), max = 0.99999)
  y_frechet <- qfrechet(u_uniform_for_frechet, loc = loc, scale = scale, shape = shape)
  
  y_mixed_frechet <- (1 - B) * y_norm + B * y_frechet
  
  ## uniform variable for A-R decision
  u_uniform <- runif(1, min = 0, max = 1)

  indicator_acceptance_rejection <- density_approx(
  t = y_mixed_frechet, h = min(h, min(diff(support_density)) / 2), t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
) / (C * density_approx(
      t = y_mixed_frechet, h = min(h, min(diff(support_density)) / 2), t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
    )
  )
  
  if (is.na(indicator_acceptance_rejection))
    {browser()}
  
  if (u_uniform <= indicator_acceptance_rejection) {
    X[k] <- y_mixed_frechet
    k <- k + 1
  }
}

```

```{r echo=FALSE, eval = TRUE}
summary(X)
hist(X)

plot(ecdf(X))
```

#### using the analytical densities
```{r echo=FALSE, eval = TRUE}

n_support_approx <- 10^3
n_min_approx <- t_min - 1 / n_support
n_max_approx <- 10^3
support_density_approx <- seq(from = n_min_approx, to = n_max_approx, by = 1 / n_support_approx)


## scaling factor (lower tail is the same for both distributions)
C_lower_tail <- 1

## the frechet distribution has a heavier tail than the semiexponential case
C_upper_tail <- max(
  density_approx(
  t = support_density_approx, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
) / density_approx(
      t = support_density_approx, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
  ), na.rm = TRUE
)

C <- max(C_lower_tail, C_upper_tail)

```

#### Accept-Reject using analytical densities
```{r echo=FALSE, eval = TRUE, warning = FALSE}
X <- c()
k = 1
h = 1e-5

set.seed(42)

a_r_analytical <- function(t_min, p, r, a, L, mu, sd, loc, scale, shape, max_iter, C) {
  
  X <- NA_real_
  iter <- 1
  
  while((is.na(X) & (iter < max_iter))) {
  
    iter <- iter + 1
    
    ## sample a binomial rv to decide whether we sample from lower or upper tail
    B <- rbinom(n = 1, size = 1, prob = p)
    
    if (B == 0) {
      r_norm <- rnorm(n = 1, mean = mu, sd = 1)
      y_norm <- ifelse(r_norm < t_min, r_norm, t_min)
    } else {
      y_norm <- 0
    }
    
    if (B == 1) {
      ## sample X ~ Frechet, conditional on X > 2
      u_uniform_for_frechet <- runif(1, min = evd::pfrechet(q = t_min, loc = loc, scale = scale, shape = shape), max = 1)
      
      ## do not allow runif == 1
      while(u_uniform_for_frechet == 1) {
        u_uniform_for_frechet <- runif(n = 1, min = evd::pfrechet(q = t_min, loc = loc, scale = scale, shape = shape), max = 1)
      }
      y_frechet <- evd::qfrechet(p = u_uniform_for_frechet, loc = loc, scale = scale, shape = shape)
    }
    
    y_mixed_frechet <- (1 - B) * y_norm + B * y_frechet
    
    ## uniform variable for A-R decision
    u_uniform <- runif(n = 1, min = 0, max = 1)
  
    indicator_acceptance_rejection <- density_analytical(
    t = y_mixed_frechet, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
  ) / (C * density_analytical(
        t = y_mixed_frechet, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
      )
    )
    
    if (u_uniform <= indicator_acceptance_rejection) {
      X <- y_mixed_frechet
      
      return(X)
    }
  }
  
  return(X) ## in this case X = NA_real_
}

X <- c()
for (i in 1:10^3) {
  X <- c(X, a_r_analytical(t_min = t_min, p = p_mixture, r = r, a = a, L = L, mu = mu, sd = sd, loc = loc, scale = scale, shape = shape, max_iter = max_iter, C = C))
}
```

```{r echo=FALSE, eval = TRUE}
summary(X)
hist(X)

plot(ecdf(X), verticals = TRUE, do.points = FALSE)
```


```{r echo=FALSE, eval = TRUE}

## reproduction distribution (conditioned on >= 1)
rgeom_greater_0 <- function(n = 1, prob) {
  
  ## sample from a Geometrix variable conditioned that it is greater than 0
  qgeom(runif(n = n, pgeom(q = 0, prob = prob),1), prob = prob)
  
}

tree_indexed_random_walk <- function(dist = rgeom_greater_0, prob, generations = 5, initial_population = 1, 
                 t_min, p_mixture, r, a, L, mu, sd, loc, scale, shape, seed = 42, max_iter = 10^5, C) {
  
  set.seed(seed)
  
  id <- 1
  label <- a_r_analytical(t_min = t_min, p = p_mixture, r = r, a = a, L = L, mu = mu, sd = sd, loc = loc, scale = scale, shape = shape, max_iter = max_iter, C = C)
    
  Z <- list("id" = id, "parent_id" = NA_real_, "generation" = 0, "parent_generation" = 0, "label" = label, "S_v" = label, "S_v_normalized" = label / psi(n = 0,r = r), "parent_S_v_normalized" = NA_real_)
  
  for (n in 1:generations) {
    
    ## get the id's of the particles in the previous generation
    particles_in_previous_generation <- Z$id[which(Z[["generation"]] == n - 1)]
    
    for (k in seq_along(particles_in_previous_generation)) {
      
      ## sample the number of offspring of the k-th particle in the last generation
      number_of_offspring_temp <- dist(n = 1, prob = prob)
      
      for (j in 1:number_of_offspring_temp) {
        
        ## sample a new label
        label_temp <- a_r_analytical(t_min = t_min, p = p_mixture, r = r, a = a, L = L, mu = mu, sd = sd, loc = loc, scale = scale, shape = shape, max_iter = max_iter, C = C)
        
        ## the new spatial position is given by the previous particle position + the new lable
        S_v_temp <- Z$S_v[particles_in_previous_generation[k]] + label_temp
        S_v_normalized_temp <- S_v_temp / psi(n = n, r = r)
        
        id <- id + 1
        Z$id <- c(Z$id, id)
        Z$parent_id <- c(Z$parent_id, particles_in_previous_generation[k])
        Z$generation <- c(Z$generation, n)
        Z$parent_generation <- c(Z$parent_generation, n - 1)
        Z$label <- c(Z$label, label_temp)
        Z$S_v <- c(Z$S_v, S_v_temp)
        Z$S_v_normalized <- c(Z$S_v_normalized, S_v_normalized_temp)
        Z$parent_S_v_normalized <- c(Z$parent_S_v_normalized, Z$S_v[particles_in_previous_generation[k]] / psi(n = n - 1, r = r))
        
      }
      
    }
  }
  
  return(Z)
}

```

```{r echo=FALSE, eval = TRUE}

prob_geom = 0.65
Z <- tree_indexed_random_walk(dist = rgeom_greater_0, prob = prob_geom, generations = 5, initial_population = 1, 
                 t_min = t_min, p_mixture = p, r = r, a = a, L = L, mu = mu, sd = sd, loc = loc, scale = scale, shape = shape, max_iter = 10^5, C = C) 

## install.packages("igraph")
library(igraph)

Z_df <- data.frame(
  id = Z$id,
  parent_id = Z$parent_id,
  generation = Z$generation,
  label = Z$label,
  S_v = Z$S_v
)

Z_df$label[which(id == 1)] <- 1

edges <- na.omit(Z_df[, c("parent_id", "id")])
g <- graph_from_data_frame(edges, directed = TRUE)



plot.igraph(g, layout = layout_as_tree, vertex.label = round(Z_df$label[which(Z_df$id == V(g)$name)], 1), vertex.label.cex = 0.85, vertex.size = 25, arrow.color = "black", edge.arrow.size = 0, vertex.color = "lightgray", main = "Galton-Watson Process Tree")

```



```{r echo=FALSE, eval = TRUE}

#plot(NULL, xlim = range(Z_df$generation), ylim = range(Z_df$S_v), 
#     xlab = "Generation", ylab = "tree-indexed random walk", 
#     xaxt = 'n', yaxt = 'n', bty = 'n')

# Add custom x and y axes
#axis(1, at = unique(Z_df$generation))
#axis(2, at = pretty(range(Z_df$S_v)))

# 0.6, 15 is gut 
prob_geom <- 0.7
n_generations <- 10

Z <- tree_indexed_random_walk(dist = rgeom_greater_0, prob = prob_geom, generations = n_generations, initial_population = 1, 
                 t_min = t_min, p_mixture = p, r = r, a = a, L = L, mu = mu, sd = sd, loc = loc, scale = scale, shape = shape, max_iter = 10^5, C = C) 

## install.packages("igraph")
library(igraph)

Z_df <- data.frame(
  id = Z$id,
  parent_id = Z$parent_id,
  generation = Z$generation,
  parent_generation = Z$parent_generation,
  label = Z$label,
  S_v = Z$S_v,
  S_v_normalized = Z$S_v_normalized,
  parent_S_v_normalized = Z$parent_S_v_normalized
)

edges <- na.omit(Z_df[, c("parent_id", "id")])
g <- graph_from_data_frame(edges, directed = TRUE)


layout_brw <- matrix(ncol = 2, nrow = vcount(g))
layout_brw[,1] <- Z_df$generation # x-axis: generation
layout_brw[,2] <- Z_df$S_v_normalized # y-axis: label

plot.igraph(g, layout = layout_brw, vertex.label = NA, vertex.size = 3, edge.arrow.size = 0, vertex.color = "lightgray")

```

```{r echo=FALSE, eval = TRUE}

## plot that shows the maximum is driven by few large increments
# r <- 0.6 
# seed = 8 
# t_min <- 10
# prob_geom <- 0.6
# n_generations <- 15

## plot that shows trend decline to the limiting speed
# r <- 0.6
# seed = 8 
# t_min <- 10
# prob_geom <- 0.75
# n_generations <- 30

r <- 0.5
seed = 8
t_min <- 2
prob_geom <- 0.875
m <- mean(rgeom_greater_0(n = 10^4, prob = prob_geom))

speed <- log(m)^(1 / r)

n_generations <- 50

Z <- tree_indexed_random_walk(dist = rgeom_greater_0, prob = prob_geom, generations = n_generations, initial_population = 1, 
                 t_min = t_min, p_mixture = p, r = r, a = a, L = L, mu = mu, sd = sd, loc = loc, scale = scale, shape = shape, seed = seed, max_iter = 10^5, C = C) 

## install.packages("igraph")
library(igraph)

Z_df <- data.frame(
  id = Z$id,
  parent_id = Z$parent_id,
  generation = Z$generation,
  parent_generation = Z$parent_generation,
  label = Z$label,
  S_v = Z$S_v,
  S_v_normalized = Z$S_v_normalized,
  parent_S_v_normalized = Z$parent_S_v_normalized
)

edges <- na.omit(Z_df[, c("parent_id", "id")])
g <- graph_from_data_frame(edges, directed = TRUE)

layout_brw <- matrix(ncol = 2, nrow = vcount(g))
layout_brw[,1] <- Z_df$generation # x-axis: generation
layout_brw[,2] <- Z_df$S_v_normalized # y-axis: label


library(ggplot2)

ggplot() +
    geom_hline(yintercept = speed, linetype = "dashed", color = "red") +
    geom_segment(data=Z_df[Z_df$id != 1, ],aes(x = parent_generation, xend = generation, y = parent_S_v_normalized, yend = S_v_normalized),color="black") +
    geom_point(data = Z_df, aes(x = generation,y = S_v_normalized), size = 2.25, colour="black") +  # adds a black border around the nodes
    geom_point(data = Z_df, aes(x = generation,y = S_v_normalized),size=1.35,colour="lightgrey") +
    scale_x_continuous(expand = c(0,1))+
    scale_y_continuous(expand = c(0,1))+ 
    theme_bw() 

# library(magrittr)
Z_df_max <- Z_df
Z_df_max <- Z_df_max %>%
  dplyr::group_by(generation) %>%
  dplyr::filter(S_v_normalized == max(S_v_normalized)) %>%
  dplyr::ungroup()

ggplot() +
    geom_hline(yintercept = speed, linetype = "dashed", color = "red") +
    geom_point(data=Z_df_max[Z_df_max$id != 1, ],aes(x = generation, y = S_v_normalized),color="black") +
    scale_x_continuous(expand = c(0,1)) +
    scale_y_continuous(expand = c(0,1)) +
    theme_classic()

## Frage: welchen Einfluss haben die Konstanten die wir haben verschwinden lassen 
#### -> Konstante verschwunden durch X | X > 2 für upper tail

```