---
title: "Simulation of tree-indexed random walks"
output: html_document
date: "2024-07-01"
---

```{r echo=FALSE, eval = FALSE}
library(evd)
```

#### define a, L, psi
```{r echo=FALSE, eval = FALSE}

# Example 2 of Gantert
psi <- function(n, r = 0.5) {
  psi <- ( r*n / log(n) )^(1 / r)
  
  return(psi)
}

L <- function(n) {
  L <- log(n)
  
  return(L)
}

# assume a = 1 (for simplification)
a <- function(n) {
  return(1)
}

r = 0.5

```

```{r echo=FALSE, eval = FALSE}
mu = 0
sd = 1

loc = 0
scale = 1
shape = 1

p = 0.75
```


#### define the support
```{r echo=FALSE, eval = FALSE}
n_min = 10^1
n_max = 10^3

support <- seq(from = n_min, to = n_max, by = 1)
```


#### plot psi vs. n
```{r echo=FALSE, eval = TRUE}
plot(support, psi(support))
lines(support, support)
```


#### semiexponential tails
```{r echo=FALSE, eval = TRUE}
n_min_tail = 10^4
n_max_tail = 10^6
N = 10^4

tail_support <- seq(from = n_min_tail, to = n_max_tail, by = (n_max_tail - n_min_tail) / N)

upper_tail_dist <- function(t, t_min, r, a, L) {
  
  prob = a(t) * exp(-L(t) * t^r) * (t >= t_min)
    
  return(prob)
}

distribution_stretched_exp_tail <- function(t, t_min, p = 0.75, r = 0.5, a = 1, L, lower_tail_dist, mu = 0, sd = 1){
  
  prob_upper_tail <- (1 - p) + p * (1 - upper_tail_dist(t = t, t_min = t_min, r = r, a = a, L = L))
  
  if (lower_tail_dist == "normal") {
      prob_lower_tail <- (1 - p) * pnorm(t, mean = mu, sd = sd)
  }
  
  prob <- ifelse(t > t_min, prob_upper_tail, ifelse(t == t_min, 1 - p, prob_lower_tail))
  
  return(prob)
}

t_min = 2
distribution_stretched_exp_tail(t = t_min, t_min, r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd)
```

#### fréchet upper tail
```{r echo=FALSE, eval = TRUE}
distribution_frechet_tail <- function(t, t_min, p = 0.75, loc = 0, scale = 1, shape = 1, lower_tail_dist = "normal", mu = 0, sd = 1) {
  
  prob_upper_tail <- (1 - p) + p * pfrechet(q = t, loc = loc, scale = scale, shape = shape, lower.tail = TRUE) 
  
  if (lower_tail_dist == "normal") {
      prob_lower_tail <- (1 - p) * pnorm(t, mean = mu, sd = sd)
  }
  
  prob <- ifelse(t > t_min, prob_upper_tail, ifelse(t == t_min, 1 - p, prob_lower_tail))
  
  return(prob)
}

```

#### plot cdf
```{r echo=FALSE, eval = TRUE, warning = FALSE}
n_min = -5
n_max = 5

support <- seq(from = n_min, to = n_max, by = 0.05)

dist_stretched_exp_tail <- distribution_stretched_exp_tail(
  t = support, t_min = 2, r = 0.5, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

dist_frechet <- distribution_frechet_tail(t = support, t_min, p = p, loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd)

plot(support, dist_stretched_exp_tail, type = "s", ylab = "distribution", xlab = "")
lines(support, pnorm(support, mean = 0, sd = 1), col = "blue")
lines(support, dist_frechet, type = "s", col = "orange")
lines(support, rep(x = 1 - p, times = length(support)), col = "lightgray", lty = 2)

```

#### approximate the density of the stretched exponential distribution
```{r echo=FALSE, eval = TRUE}
n_support <- 10^3
n_min <- -10 + 1 / n_support
n_max <- 10
support_density <- seq(from = n_min, to = n_max, by = 1 / n_support)

## use a centered difference approximation to estimate the density
density_approx <- function(
  t, h = 1e-5, t_min = 2, p = 0.75, upper_tail_dist = "stretched", r = 0.5, a = function() {1}, L = log, loc = 0, scale = 1, shape = 1, lower_tail_dist = "normal", mu = 0, sd = 1
) {
  
  if (h >= min(diff(support_density))) {
    warning("The numerical differentiation will not work at the discontinuity t_min.")
  }
  
  if (upper_tail_dist == "stretched") {
    density_approx <- ( 
      distribution_stretched_exp_tail(
        t = t+h, t_min = t_min, p = p, r = r, a = a, L = L, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      ) - distribution_stretched_exp_tail(
        t = t-h, t_min = t_min, p = p, r = r, a = a, L = L, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      ) 
    ) / (2 * h)
  } else if (upper_tail_dist == "frechet") {
    density_approx <- ( 
      distribution_frechet_tail(
        t = t+h, t_min, p = p, loc = loc, scale = scale, shape = shape, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      ) - distribution_frechet_tail(
          t = t-h, t_min, p = p, loc = loc, scale = scale, shape = shape, lower_tail_dist = lower_tail_dist, mu = mu, sd = sd
      )
    ) / (2 * h)
  }
  
  ## density is not defined at the discontinuity t = t_min
  density_approx <- ifelse(t != t_min, density_approx, NA_real_)
  
  return(density_approx)
}

```

#### analytical representation of density
```{r echo=FALSE, eval = TRUE}
density_analytical <- function(
    t, t_min = 2, p = 0.75, upper_tail_dist = "stretched", r = 0.5, a = function() {1}, L = log, loc = 0, scale = 1, shape = 1, lower_tail_dist = "normal", mu = 0, sd = 1
) {
  
  if (lower_tail_dist == "normal") {
    density_lower_tail <- (1 - p) * dnorm(t, mean = mu, sd = sd)
  }
  
  if (upper_tail_dist == "stretched") {
    density_upper_tail <- ifelse(t > 0, p * ( a(t) * t^(r-1) * exp(-L(t)*t^r) * (r * log(t) + 1) ), NA_real_)
  } else if (upper_tail_dist == "frechet") {
      density_upper_tail <- ifelse(t > 0, p * dfrechet(t, loc = loc, scale = scale, shape = shape), NA_real_)
  }
  
  density <- ifelse(t < t_min, density_lower_tail, density_upper_tail)
  
  return(density)
}

```

#### plot pdf
```{r echo=FALSE, eval = TRUE, warning = FALSE}

h = min(1e-7, min(diff(support_density)) / 2)

dist_stretched_exp_tail_approx <- distribution_stretched_exp_tail(
  t = support_density, t_min = 2, r = 0.5, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_stretched_exp_tail_approx <- density_approx(
  t = support_density, h = h, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_stretched_exp_tail_analytical <- density_analytical(
  t = support_density, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_frechet_tail_approx <- density_approx(
  t = support_density, h = h, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_frechet_tail_analytical <- density_analytical(
  t = support_density, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
)

density_norm <- dnorm(support_density, mean = mu, sd = sd)

y_max = 0.075 + max(
  max(density_stretched_exp_tail, na.rm = TRUE), 
  max(density_frechet, na.rm = TRUE), 
  max(density_norm)
)

plot(support_density, density_stretched_exp_tail_approx, type = "s", ylim = c(0, y_max), ylab = "density", xlab = "")
lines(support_density, density_stretched_exp_tail_analytical, col = "blue", lty = 2)

plot(support_density, density_frechet_tail_approx, type = "s", ylim = c(0, y_max), ylab = "density", xlab = "")
lines(support_density, density_frechet_tail_analytical, col = "red", lty = 2)

```




#### Acceptance-Rejection Algorithm

#### using approximated densities
```{r echo=FALSE, eval = TRUE}

n_support_approx <- 10^3
n_min_approx <- t_min - 1 / n_support
n_max_approx <- 10^3
support_density_approx <- seq(from = n_min_approx, to = n_max_approx, by = 1 / n_support_approx)


## scaling factor (lower tail is the same for both distributions)
C_lower_tail <- 1

## the frechet distribution has a heavier tail than the semiexponential case
C_upper_tail <- max(
  density_approx(
  t = support_density_approx, h = min(1e-7, min(diff(support_density_approx)) / 2), t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
) / density_approx(
      t = support_density_approx, h = min(1e-7, min(diff(support_density_approx)) / 2), t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
  ), na.rm = TRUE
)

C <- max(C_lower_tail, C_upper_tail)

```

#### Accept-Reject using the numerical approximation of densities
```{r echo=FALSE, eval = TRUE, warning = FALSE}
X <- c()
k = 1
h = 1e-5

set.seed(42)

while (k < 5 * 10^3) {
  B <- rbinom(1, 1, prob = p)
  y_norm <- min(rnorm(n = 1, mean = mu, sd = 1), t_min - 2*h)
  
  ## sample X ~ Frechet, conditional on X > 2
  u_uniform_for_frechet <- runif(1, min = pfrechet(2, loc = loc, scale = scale, shape = shape), max = 0.99999)
  y_frechet <- qfrechet(u_uniform_for_frechet, loc = loc, scale = scale, shape = shape)
  
  y_mixed_frechet <- (1 - B) * y_norm + B * y_frechet
  
  ## uniform variable for A-R decision
  u_uniform <- runif(1, min = 0, max = 1)

  indicator_acceptance_rejection <- density_approx(
  t = y_mixed_frechet, h = min(h, min(diff(support_density)) / 2), t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
) / (C * density_approx(
      t = y_mixed_frechet, h = min(h, min(diff(support_density)) / 2), t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
    )
  )
  
  if (is.na(indicator_acceptance_rejection))
    {browser()}
  
  if (u_uniform <= indicator_acceptance_rejection) {
    X[k] <- y_mixed_frechet
    k <- k + 1
  }
}

```

```{r echo=FALSE, eval = TRUE}
summary(X)
hist(X)

plot(ecdf(X))
```

#### using the analytical densities
```{r echo=FALSE, eval = TRUE}

n_support_approx <- 10^3
n_min_approx <- t_min - 1 / n_support
n_max_approx <- 10^3
support_density_approx <- seq(from = n_min_approx, to = n_max_approx, by = 1 / n_support_approx)


## scaling factor (lower tail is the same for both distributions)
C_lower_tail <- 1

## the frechet distribution has a heavier tail than the semiexponential case
C_upper_tail <- max(
  density_approx(
  t = support_density_approx, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
) / density_approx(
      t = support_density_approx, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
  ), na.rm = TRUE
)

C <- max(C_lower_tail, C_upper_tail)

```

#### Accept-Reject using analytical densities
```{r echo=FALSE, eval = TRUE, warning = FALSE}
X <- c()
k = 1
h = 1e-5

set.seed(42)

while (k < 10^3) {
  B <- rbinom(1, 1, prob = p)
  y_norm <- min(rnorm(n = 1, mean = mu, sd = 1), t_min - 2*h)
  
  ## sample X ~ Frechet, conditional on X > 2
  u_uniform_for_frechet <- runif(1, min = pfrechet(2, loc = loc, scale = scale, shape = shape), max = 1)
  while(u_uniform_for_frechet == 1) {
    u_uniform_for_frechet <- runif(1, min = pfrechet(2, loc = loc, scale = scale, shape = shape), max = 1)
  }
  y_frechet <- qfrechet(u_uniform_for_frechet, loc = loc, scale = scale, shape = shape)
  
  y_mixed_frechet <- (1 - B) * y_norm + B * y_frechet
  
  ## uniform variable for A-R decision
  u_uniform <- runif(1, min = 0, max = 1)

  indicator_acceptance_rejection <- density_analytical(
  t = y_mixed_frechet, t_min = t_min, p = p, upper_tail_dist = "stretched", r = r, a = a, L = L, lower_tail_dist = "normal", mu = mu, sd = sd
) / (C * density_analytical(
      t = y_mixed_frechet, t_min = t_min, p = p, upper_tail_dist  = "frechet", loc = loc, scale = scale, shape = shape, lower_tail_dist = "normal", mu = mu, sd = sd
    )
  )
  
  if (is.na(indicator_acceptance_rejection))
    {browser()}
  
  if (u_uniform <= indicator_acceptance_rejection) {
    X[k] <- y_mixed_frechet
    k <- k + 1
  }
}

```

```{r echo=FALSE, eval = TRUE}
summary(X)
hist(X)

plot(ecdf(X), verticals = TRUE, do.points = FALSE)
```